Hybrid MILP to efficiently and accuratly \\ solve hard DNN verification instances

### 2024-09-24 ICLR25/formula.tex
update

Let $z$ be a neuron of the DNN we want to evaluate, for instance computing an upper bound on the value it can take. We denote by $n< z$ the fact that a neuron appears on a layer before the layer of $z$, and $n \leq z$ if it can also be $z$.
Let us denote (Sol\_max$_X^z(n))_{n \leq z}$ is a solution of $\mathcal{M}_X$ 
maximizing $z$. In particular, Sol\_max$_X^z(z)$ is the maximum of $z$ under $\mathcal{M}_X$.

Consider (Sol\_max$_\emptyset^z(n))_{n \leq z}$ (this is obtained by {\em one} call to an LP solver, so can be obtained very efficiently).
An accurate utility function is to evaluate 
$\Improve\_max^z(a)=$ Sol\_max$_\emptyset^z(z) -$ Sol\_max$_{\{a\}}^z(z)$, 
representing how much opening neuron $a < z$ reduces the maximum computed for $z$
compared with using only LP. 
We have $\Improve\_max^z(a)\geq 0$ as Sol\_max$_{\{a\}}^z$ fullfils all the constraints of 
$\mathcal{M}_\emptyset$, so Sol\_max$_{\{a\}}^z(z) \leq$ Sol\_max$_\emptyset^z(z)$.
Similarly, we define (Sol\_min$_\emptyset^z(n))_{n \leq z}$ and 
$\Improve\_min_a^z$. Calling MILP on $\mathcal{M}_{\{a\}}$ for each neuron $a \leq z$
would however be very time consuming when the number of neurons $a$ to evaluate is large.
Instead, we focus on the following upper bounding of $\Improve\_max_a^z$.

@@

For all neurons $n$, let $sol(n)=$Sol\_max$_\emptyset^z(n)$ be the value of neuron $a$
in the solution of the LP instance Sol\_max$_\emptyset^z$ to maximize $z$.
%For one layer before the target layer, the formula is simple and most accurate. 
To estimate $\Improve\_max_a^z$, we define $\Utility$\_max$^z(a)$, first for neurons $a$ one layer before $z$:
%we first need to run $M^a_{\emptyset}$ to compute the upper bound of $a$  to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

	\begin{align}
		\Utility\_max^z(a) = |W_{az} \times (sol(\hat{a})- \ReLU(sol(a)))|.
	\end{align}

	%In particular, if $sol(\hat{a})=\ReLU(sol(a))$, then we will have 
	%\begin{align*}
%		Utility^z(a) = 0.
%	\end{align*}

@@

\begin{proposition}
$0 \leq \Improve\_max^z(a) \leq \Utility\_\max^z(a)$. 
\end{proposition}

\begin{proof}
Consider $(sol'(n))_{n \leq z})_{n \leq z})$ with
$sol'(n)=sol(n)$ for all $n \notin \{z,\hat{a}\}$.
In particular,  $sol'(a) = sol(a)$.

Now, define $sol'(\hat{a}) = \ReLU(sol(a))$. 
That is,  
sol'$(\hat{a})$ is the correct value for $\hat{a}$ (obtained if we open neuron $a$), 
compared to the LP abstraction for $sol(\hat{a})$.

Last, we define $sol'(z)=sol(z) + W_{az} \times (\ReLU(sol(a)) - sol(\hat{a})) = sol(z) - Utility\_max_z(a)$.

Now, it is easy to check that $(sol'(n))_{n \leq z}$ satisfies the constraints in 
$\mathcal{M}_{\{a\}}$, as openning $a$ changes the value of $\hat{a}$ from
$sol(\hat{a})$ to $\ReLU(sol(a))$, and the contribution from $a$ to $z$ is 
$W_{az}$. As $sol'(z)$ is a solution of $\mathcal{M}_{\{a\}}$, it is smaller or equal to the maximal solution: $sol'(z) \leq$ Sol\_max$_{\{a\}}^z(z)$. That is, 
$sol(z)-sol'(z) \geq sol(z) -$ Sol\_max$_{\{a\}}^z(z)$, i.e. 
$\Improve\_max^z(a) \leq Utility\_\max^z(a)$.
\end{proof}

Thus, Utility\_max$^z(a)$ can be used to approximate 
$\Improve\_max^z(a)$. In particular, consider a very frequent case where $W_{az}<0$.
In this case, to maximize $z$, the LP engine sets $sol(\hat{a})$ to its minimal value\dots which happens to be $sol(\hat{a})=\ReLU(a)$, thanks to Proposition \ref{LP}. 
In this frequent case, we have Utility\_max$^z(a)=0$, and thus $\Improve\_max^z(a)=0$. 
In particular, this node will have the smallest utility (thus will not get picked in the open nodes $X$), and indeed it is not having any impact on  Sol\_max$_{\{a\}}^z(z)$. 
This is one stricking difference (but not the only one) with choosing utility based on 
$|W_{az}|$ \cite{DivideAndSlide}.

%	
%	Similarly, let $sol(b)$ be the value of $b$ in the LP solution of lower bound of $a$, and $sol(\hat{b})$ be the value of $\hat{b}$. Then the formula to estimate improvement of lower bound of $b$ is: \begin{align*}
%		Improve\_min^z(b) \approx -W_{ba}(sol(\hat{b})-\ReLU(sol(b))).
%	\end{align*}

%To explain the formula, we use upper bound and the case that $W_{ba} > 0$ as an example. To compute the upper bound of $a$, $\hat{b}$ should be as large as possible. In the LP model, for fixed $sol(b)$, the upper bound of $\hat{b}$ may be larger than $\ReLU(sol(b))$. This is because in LP model, the upper bound of $sol(\hat{b})$ is decided by the linear approximation rather than $\ReLU$ function. So, when neuron $b$ is open, if $sol(b)$ do not change, then the upper bound of $a$ will be improved because the value of $sol(\hat{b})$ will be lower to $\ReLU(sol(b))$.
% 			
%Of course changing other variables may also effect the upper bound, but our experiments show that, the change from $sol(\hat{b})$ to $\ReLU(sol(b))$ is the major part of improvement.

@@

Now, consider neurons $a$ two layers before $z$. We use $b$ to denote neurons in the layer between $a$ and $z$. If we open $a$, then the change in the weight of $a$ will still be $sol(\hat{a})-\ReLU(sol(a))$ as above. Its impact on $z$ is no more direct with $W_{az}$, but it is through all the $b$ in the intermediate layer. 
Based on Proposition \ref{LP}, we can evaluate how opening $a$ impacts each value of 
$\hat{b}$, through coefficient $k(\hat{b})$, by using the upper and lower bound
$\UB(b),\LB(b)$:
\begin{itemize}
 \item If $W_{bz}\leq0$, then $k(\hat{b})=W_{ab}$ if $sol(b) > 0$, and $k(\hat{b})=0$ if $sol(b) < 0$
 \item If $W_{bz}>0$, then $k(\hat{b})=W_{ab} \frac{\UB(b)}{\UB(b)-\LB(b)}$
\end{itemize}
%$\sum_b W_{bz}W_{ab}$ to replace the coefficient $W_{az}$ in last subsection.

%To estimate $\Improve\_max_a^z$ , 
Indeed, if $W_{bz}\leq0$, then the LP solver sets $sol(\hat{b})$ to the lowest possible value to maximize $z$, which happens to be $\ReLU(b)$ according to Proposition \ref{LP}.
If $sol(b) < 0$, then we have a clip at $sol(\hat{b})=\ReLU(b)=0$ and opening $a$ does not help. If $sol(b) > 0$, then we have $sol(\hat{b})=\ReLU(b)=sol(b)$ and every change from 
$a$ is reflected with a coefficient $W_{ab}$ on $\hat{b}$. 
Last, if $W_{bz}>0$, then according to Proposition \ref{LP}, the LP solver
sets $sol(\hat{b}) = sol(b) \frac{\UB(b)}{\UB(b)-\LB(b)}$ to maximize $z$.

%We then define 
%\begin{align*}
%	Utility\_max^z(a) = (sol(\hat{a})-\ReLU(sol(a)))\sum_b k(b).
%\end{align*}

We define a difference $\Delta$ to account for the fact that a neuron $b$ 
can not exceed $\LB(b),\UB(b)$:
	\begin{align*}
		&\Delta(\hat{a}) = sol(\hat{a}) - \ReLU(sol(a)) \geq 0\\
			&\Delta(\hat{b}) =
		\begin{cases}
			sol(\hat{b}) - \min(\UB(b), sol(\hat{b}) - \Delta(\hat{a}) k(\hat{b})), & \text{if }  W_{bz} > 0 \\
			sol(\hat{b}) - \max(\LB(b), 0, sol(\hat{b}) -\Delta(\hat{a}) k(\hat{b})), & \text{if }W_{bz} < 0\\
		\end{cases}
		\end{align*}

@@

$$ \Utility\_max^z(a) = \sum_b W_{bz} \Delta(\hat{b})$$

We proceed inductively in the same way to define $\Utility\_max^z(a)$ for deeper neuron $a$.

@@

\begin{definition}\label{3layer}
Let $\UB$ and $\LB$ denote the precomputed upper bounds and lower bounds used in building MILP models. We define the following function $h$ for all neurons $b$ in two layers before the target neuron $z$ as follows:
	\begin{align}
		&v_0 = sol(\hat{b}), v_1 = \ReLU(sol(b)), v_2 = \frac{\UB(b)sol(b)-\UB(b)\LB(b)}{\UB(b)-\LB(b)}\\
		&h(b) =
		\begin{cases}
			\frac{v_0-v_1}{v_2-v_1}, & \text{if } v_2-v_1 > 0\\
			0.5, & \text{otherwise.}
		\end{cases}
	\end{align} 
\end{definition}

@@

	First, $\Delta(a) = \ReLU(sol(a))-sol(\hat{a})$.

To compute $\Delta(b)$ for neurons $b$ in two layer before $z$, we define \begin{align}
	&u_0 = \max(\LB(b),\min(\UB(b),  sol(b)+\Delta(a)W_{ab}))\\
	&u_1 = \begin{cases}
		\ReLU(u_0)+h(b)(\frac{\UB(c)u_0-\UB(b)\LB(b)}{\UB(b)-\LB(b)}-\ReLU(u_0)), & \text{if }\LB(b) < 0\\
	u_0, & \text{if }  \LB(b) \geq 0
	\end{cases}\\
	&\Delta(b) = u_1-sol(\hat{b})
\end{align}

	To compute $\Delta(c)$ for neurons $c$ in one layer before $z$, we define 
	\begin{align}
		&w_0 = \sum_b \Delta(b)W_{bc}\\
		&w_1 = \min(\UB(c),sol(c)+w_0)\\		
		&\Delta(c) =
		\begin{cases}
			w_1-sol({c}), & \text{if }W_{cz} > 0 \text{ and } \LB(c)\geq 0\\
		k(c)(w_1-sol({c})), & \text{if }W_{cz} > 0 \text{ and } \LB(c)< 0\\
		\ReLU(w_1)-sol(\hat{c})	, & \text{if }  W_{cz} < 0
		\end{cases}\\
		&\Delta(z) = \sum_c \Delta(c)W_{cz}\\
		&\Utility\_max^z(a) = -\Delta(z)
	\end{align}
\end{definition}

>>>>

Let $z$ be a neuron of the DNN we want to evaluate, for instance computing an upper bound on the value it can take. We denote by $n< z$ the fact that a neuron appears on a layer before the layer of $z$, and $n \leq z$ if it can also be $z$.
Let us denote ($\Sol\_\max_X^z(n))_{n \leq z}$ is a solution of $\mathcal{M}_X$ 
maximizing $z$. In particular, $\Sol\_\max_X^z(z)$ is the maximum of $z$ under $\mathcal{M}_X$.

Consider ($\Sol\_\max_\emptyset^z(n))_{n \leq z}$ (this is obtained by {\em one} call to an LP solver, so can be obtained very efficiently).
An accurate utility function is to evaluate 
$\Improve\_\max^z(a)=$ $\Sol\_\max_\emptyset^z(z) -$ $\Sol\_\max_{\{a\}}^z(z)$, 
representing how much opening neuron $a < z$ reduces the maximum computed for $z$
compared with using only LP. 
We have $\Improve\_\max^z(a)\geq 0$ as $\Sol\_\max_{\{a\}}^z$ fullfils all the constraints of 
$\mathcal{M}_\emptyset$, so $\Sol\_\max_{\{a\}}^z(z) \leq$ $\Sol\_\max_\emptyset^z(z)$.
Similarly, we define ($\Sol\_\min_\emptyset^z(n))_{n \leq z}$ and 
$\Improve\_\min^z(a)$. Calling MILP on $\mathcal{M}_{\{a\}}$ for each neuron $a \leq z$
would however be very time consuming when the number of neurons $a$ to evaluate is large.
Instead, we focus on the following upper bounding of $\Improve\_\max^z(a)$.

@@

For all neurons $n$, let $\sol(n)=$$\Sol\_\max_\emptyset^z(n)$ be the value of neuron $a$
in the solution of the LP instance $\Sol\_\max_\emptyset^z$ to maximize $z$.
%For one layer before the target layer, the formula is simple and most accurate. 
To estimate $\Improve\_\max^z(a)$, we define $\Utility\_\max^z(a)$, first for neurons $a$ one layer before $z$:
%we first need to run $M^a_{\emptyset}$ to compute the upper bound of $a$  to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

	\begin{align}
		\Utility\_\max\nolimits^z(a) = |W_{az} \times (\sol(\hat{a})- \ReLU(\sol(a)))|.
	\end{align}

	%In particular, if $\sol(\hat{a})=\ReLU(\sol(a))$, then we will have 
	%\begin{align*}
%		Utility^z(a) = 0.
%	\end{align*}

@@

\begin{proposition}
$0 \leq \Improve\_\max^z(a) \leq \Utility\_\max^z(a)$. 
\end{proposition}

\begin{proof}
Consider $(\sol'(n))_{n \leq z})_{n \leq z})$ with
$\sol'(n)=\sol(n)$ for all $n \notin \{z,\hat{a}\}$.
In particular,  $\sol'(a) = \sol(a)$.

Now, define $\sol'(\hat{a}) = \ReLU(\sol(a))$. 
That is,  
$\sol'(\hat{a})$ is the correct value for $\hat{a}$ (obtained if we open neuron $a$), 
compared to the LP abstraction for $\sol(\hat{a})$.

Last, we define $\sol'(z)=\sol(z) + W_{az} \times (\ReLU(\sol(a)) - \sol(\hat{a})) = \sol(z) - \Utility\_\max^z(a)$.

Now, it is easy to check that $(\sol'(n))_{n \leq z}$ satisfies the constraints in 
$\mathcal{M}_{\{a\}}$, as opening $a$ changes the value of $\hat{a}$ from
$\sol(\hat{a})$ to $\ReLU(\sol(a))$, and the contribution from $a$ to $z$ is 
$W_{az}$. As $\sol'(z)$ is a solution of $\mathcal{M}_{\{a\}}$, it is smaller or equal to the maximal solution: $\sol'(z) \leq$ $\Sol\_\max_{\{a\}}^z(z)$. That is, 
$\sol(z)-\sol'(z) \geq \sol(z) -$ $\Sol\_\max_{\{a\}}^z(z)$, i.e. 
$\Improve\_\max^z(a) \leq \Utility\_\max^z(a)$.
\end{proof}

Thus, $\Utility\_\max^z(a)$ can be used to approximate 
$\Improve\_\max^z(a)$. In particular, consider a very frequent case where $W_{az}<0$.
In this case, to maximize $z$, the LP engine sets $\sol(\hat{a})$ to its minimal value\dots which happens to be $\sol(\hat{a})=\ReLU(a)$, thanks to Proposition \ref{LP}. 
In this frequent case, we have $\Utility\_\max^z(a)=0$, and thus $\Improve\_\max^z(a)=0$. 
In particular, this node will have the smallest utility (thus will not get picked in the open nodes $X$), and indeed it is not having any impact on  $\Sol\_\max_{\{a\}}^z(z)$. 
This is one stricking difference (but not the only one) with choosing utility based on 
$|W_{az}|$ \cite{DivideAndSlide}.

%	
%	Similarly, let $\sol(b)$ be the value of $b$ in the LP solution of lower bound of $a$, and $\sol(\hat{b})$ be the value of $\hat{b}$. Then the formula to estimate improvement of lower bound of $b$ is: \begin{align*}
%		Improve\_min^z(b) \approx -W_{ba}(\sol(\hat{b})-\ReLU(\sol(b))).
%	\end{align*}

%To explain the formula, we use upper bound and the case that $W_{ba} > 0$ as an example. To compute the upper bound of $a$, $\hat{b}$ should be as large as possible. In the LP model, for fixed $\sol(b)$, the upper bound of $\hat{b}$ may be larger than $\ReLU(\sol(b))$. This is because in LP model, the upper bound of $\sol(\hat{b})$ is decided by the linear approximation rather than $\ReLU$ function. So, when neuron $b$ is open, if $\sol(b)$ do not change, then the upper bound of $a$ will be improved because the value of $\sol(\hat{b})$ will be lower to $\ReLU(\sol(b))$.
% 			
%Of course changing other variables may also effect the upper bound, but our experiments show that, the change from $\sol(\hat{b})$ to $\ReLU(\sol(b))$ is the major part of improvement.

@@

Now, consider neurons $a$ two layers before $z$. We use $b$ to denote neurons in the layer between $a$ and $z$. If we open $a$, then the change in the weight of $a$ will still be $\sol(\hat{a})-\ReLU(\sol(a))$ as above. Its impact on $z$ is no more direct with $W_{az}$, but it is through all the $b$ in the intermediate layer. 
Based on Proposition \ref{LP}, we can evaluate how opening $a$ impacts each value of 
$\hat{b}$, through coefficient $k(\hat{b})$, by using the upper and lower bound
$\UB(b),\LB(b)$:
\begin{itemize}
 \item If $W_{bz}\leq0$, then $k(\hat{b})=W_{ab}$ if $\sol(b) > 0$, and $k(\hat{b})=0$ if $\sol(b) < 0$
 \item If $W_{bz}>0$, then $k(\hat{b})=W_{ab} \frac{\UB(b)}{\UB(b)-\LB(b)}$
\end{itemize}
%$\sum_b W_{bz}W_{ab}$ to replace the coefficient $W_{az}$ in last subsection.

%To estimate $\Improve\_\max_a^z$ , 
Indeed, if $W_{bz}\leq0$, then the LP solver sets $\sol(\hat{b})$ to the lowest possible value to maximize $z$, which happens to be $\ReLU(b)$ according to Proposition \ref{LP}.
If $\sol(b) < 0$, then we have a clip at $\sol(\hat{b})=\ReLU(b)=0$ and opening $a$ does not help. If $\sol(b) > 0$, then we have $\sol(\hat{b})=\ReLU(b)=\sol(b)$ and every change from 
$a$ is reflected with a coefficient $W_{ab}$ on $\hat{b}$. 
Last, if $W_{bz}>0$, then according to Proposition \ref{LP}, the LP solver
sets $\sol(\hat{b}) = \sol(b) \frac{\UB(b)}{\UB(b)-\LB(b)}$ to maximize $z$.

%We then define 
%\begin{align*}
%	Utility\_max^z(a) = (\sol(\hat{a})-\ReLU(\sol(a)))\sum_b k(b).
%\end{align*}

We define a difference $\Delta$ to account for the fact that a neuron $b$ 
can not exceed $\LB(b),\UB(b)$:
	\begin{align*}
		&\Delta(\hat{a}) = \sol(\hat{a}) - \ReLU(\sol(a)) \geq 0\\
			&\Delta(\hat{b}) =
		\begin{cases}
			\sol(\hat{b}) - \min(\UB(b), \sol(\hat{b}) - \Delta(\hat{a}) k(\hat{b})), & \text{if }  W_{bz} > 0 \\
			\sol(\hat{b}) - \max(\LB(b), 0, \sol(\hat{b}) -\Delta(\hat{a}) k(\hat{b})), & \text{if }W_{bz} < 0\\
		\end{cases}
		\end{align*}

@@

$$ \Utility\_\max\nolimits^z(a) = \sum_b W_{bz} \Delta(\hat{b})$$

We proceed inductively in the same way to define $\Utility\_\max^z(a)$ for deeper neuron $a$.

@@

\begin{definition}\label{3layer}
Let $\UB$ and $\LB$ denote the precomputed upper bounds and lower bounds used in building MILP models. We define the following function $h$ for all neurons $b$ in two layers before the target neuron $z$ as follows:
	\begin{align}
		&v_0 = \sol(\hat{b}), v_1 = \ReLU(\sol(b)), v_2 = \frac{\UB(b)\sol(b)-\UB(b)\LB(b)}{\UB(b)-\LB(b)}\\
		&h(b) =
		\begin{cases}
			\frac{v_0-v_1}{v_2-v_1}, & \text{if } v_2-v_1 > 0\\
			0.5, & \text{otherwise.}
		\end{cases}
	\end{align} 
\end{definition}

@@

	First, $\Delta(a) = \ReLU(\sol(a))-\sol(\hat{a})$.

To compute $\Delta(b)$ for neurons $b$ in two layer before $z$, we define \begin{align}
	&u_0 = \max(\LB(b),\min(\UB(b),  \sol(b)+\Delta(a)W_{ab}))\\
	&u_1 = \begin{cases}
		\ReLU(u_0)+h(b)(\frac{\UB(c)u_0-\UB(b)\LB(b)}{\UB(b)-\LB(b)}-\ReLU(u_0)), & \text{if }\LB(b) < 0\\
	u_0, & \text{if }  \LB(b) \geq 0
	\end{cases}\\
	&\Delta(b) = u_1-\sol(\hat{b})
\end{align}

	To compute $\Delta(c)$ for neurons $c$ in one layer before $z$, we define 
	\begin{align}
		&w_0 = \sum_b \Delta(b)W_{bc}\\
		&w_1 = \min(\UB(c),\sol(c)+w_0)\\		
		&\Delta(c) =
		\begin{cases}
			w_1-\sol({c}), & \text{if }W_{cz} > 0 \text{ and } \LB(c)\geq 0\\
		k(c)(w_1-\sol({c})), & \text{if }W_{cz} > 0 \text{ and } \LB(c)< 0\\
		\ReLU(w_1)-\sol(\hat{c})	, & \text{if }  W_{cz} < 0
		\end{cases}\\
		&\Delta(z) = \sum_c \Delta(c)W_{cz}\\
		&\Utility\_\max^z(a) = -\Delta(z)
	\end{align}
\end{definition}
